---
title: "Outline (Inclass notes)"
output: html_document
date: "2023-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

### 0. Load the `tidyverse` Packages

This section loads the packages we need in this lecture.

```{r}
library(tidyverse)
```

### 1. Import and Export the V-Dem Data

This section loads the VDEM dataset and describe its basic information

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

### 2. to take a subset of *columns* in the existing data: `select`

Our second step is to select economic development indicators

First we look at the identifiers of the indicator

```{r}
names(d)
```

Select country year column --\> only distinct

Select country year column --\> only distinct country_names

```{r}
d |> select(country_name, country_id, year) |>
  distinct(country_name)
```

Select only e_gdp (GDP) and e_gdppc (GDP per capita)

```{r}
d |> select(country_name, country_id, year, e_gdp, e_gdppc)
```

Create a new dataset

```{r}
d_gdp <- d |>
  select(country_name, country_id, year, e_gdp, e_gdppc)
```

### 3. to rename columns: `rename`

\*does not make any changes to the original object in the environment (d) --\> just rename d_gdp

```{r}
d_gdp <- d_gdp |>
  rename("GDP" = "e_gdp","GDP_per_Capita" = "e_gdppc", "Country" = "country_name", "ID" = "country_id", "Year" = "year")
```

if you do not do the assignment, there will be no changes to the object in the environment (e.g. d_gdp2) --\> you can open a new object in order not to overwrite the original dataset

\*X need to create new object if you want to view vs if you want to **reuse** the object

```{r}
d_gdp2 <- d
d_gdp2 |>
  rename("GDP" = "e_gdp","GDP_per_Capita" = "e_gdppc", "Country" = "country_name", "ID" = "country_id", "Year" = "year")
d_gdp2 #no changes should be shown
```

### 4. to take a subset of *rows* by some simple conditions: `slice_`

get countries-years with highest GDP

```{r}
d_gdp |>
  slice_max(order_by = GDP, n = 10)
```

get random sample (every time you run, you get a different result)

```{r}
d_gdp |>
  slice_sample(n = 10)

```

### 5. to take a subset of *rows* by some more complicated conditions: `filter`

want 2000-2005

```{r}
d_gdp |>
  filter(Year>=2000 & Year<=2005 & Country == "China")
```

### 6. to sort the rows based on the value of one or multiple columns: `arrange`

arrange by GDP per Capita (asc.)

```{r}
d_gdp |>
  arrange(GDP_per_Capita)
```

arrange by GDP per Capita (desc.) use minus "-" sign

```{r}
d_gdp |>
  arrange(-GDP_per_Capita)
```

### 7. to perform (4) (5) (6) group by group: `group_by`, `ungroup`

create a dataset that contains the year of the highest development level for each country/region respectively

1.  perform a data availability/ integrity check --\> aggregate the data into new country-levle data set
    1.  average development level form 1984 to 2022
    2.  magnitute of growth from 1984 to 2022

```{r}
d_gdp |>
  group_by(Country) |>
  slice_max(GDP, n = 1)
```

compared with w/o group by --\> only show one record (max val of ALL) instead of max val of EACH

```{r}
#how many countries are there for each country?
d_gdp |>
  group_by(Country) |>
  count() #count number of rows for each country
  
```

```{r}
#country when year has the worst GDP
d_gdp |>
  group_by(Country) |>
  slice_min(order_by = GDP, n = 1)
```

### 8. to create new columns in the data: `group_by`, `mutate`, `ungroup`

**Mutate function**

create new columns

```{r}
#create new column "new"
d_gdp |>
  mutate(new = 1)

#create new column as function of old column
d_gdp |>
  mutate(GDP2 = GDP)

#add columns with new operations
d_gdp |>
  mutate(GDP3 = log(GDP))
```

create columns based on characteristics of a group

```{r}
#new column = GDP relative to the average GDP in the world
d_gdp |>
  mutate(relative_GDP = GDP/mean(GDP, na.rm = TRUE)) #use na.rm to ignore missing values, else NA output
```

```{r}
#new column = the GDP relative to the average of the country in the world 1984 to 2022
d_gdp |>
  group_by(Country)|>
  mutate(relative_country_GDP = GDP/(mean(GDP, na.rm = TRUE)))
```

```{r}
#country-year development with reference to that of 1984
d_gdp |>
  group_by(Country) |>
  arrange(Year) |>
  mutate(GDP_over_1984 = GDP/first(GDP)) |>
  ungroup() |>
  arrange(Country, Year)
```

```{r}
#year-on-year development level with reference tot htat of 1984
d_gdp |>
  group_by(Country) |>
  mutate(GDP_yoy = GDP - lag(GDP, n = 1)) |>
  ungroup() |>
  arrange(Country, Year)

```

### 9. to summarize the data: `group_by`, `summarise`, `ungroup`

\*we seldom use summarise to get summary statistics for the whole table

```{r}
#average GDP level of the world
d_gdp |>
  summarize(gdp_average = mean(GDP, na.rm = TRUE), gdp_average_per_capita = mean(GDP_per_Capita, na.rm = TRUE))
```

```{r}
#average development level for each country
d_gdp |>
  group_by(Country) |>
  summarize(gdp_average = mean(GDP, na.rm = TRUE))
```

```{r}
#GDP growth and GDP per capira growth 2019 over 1984
d_gdp |>
  #filter years >= 1984, years <= 2019
  filter(Year >= 1984 & Year <= 2019) |>
  group_by(Country) |>
  arrange(Year) |>
  summarize(gdp_growth = (last(GDP) - first(GDP))/last(GDP), gdp_growth_per_capita = (last(GDP_per_Capita) - first(GDP_per_Capita))/last(GDP_per_Capita)) |>
  ungroup() |>
  arrange(Country)
```

\*we want to see what data is available in a data set and see if there are some missing data points

```{r}
#data integrity check **for checking availability
d_gdp |>
  group_by(Country) |>
  #create a column to indicate missing GDP
  mutate (gdp_missing = is.na(GDP)) |>
  summarize(n_gdp_missing = sum(gdp_missing))
```

\*\* use of Pipe "\|\>"

-   subset of variables and filter down to show the step-by-step flow and reduce operators

-   consequence: multiple steps -\> need to product some intermediary outputs
