---
title: "Lecture 10 Text Mining: Text Wrangling: Basic String Operations"
author: "Wong Nicole"
date: "2023-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

-   Rising popularity of LLM

    -   Access data from internet websites

    -   Clean data (irrelevant parts from extractions) using basic string operations

-   How to wrangle text data using the `stringr` element of Tidyverse.

## Load Packages

```{r, message=FALSE}
library(tidyverse) # Package for data work

library(knitr) # Package for data presentation at Rmarkdown
library(kableExtra)  # Package for data presentation at Rmarkdown
```

## Load Data

```{r, results="asis"}
d = read_rds("Lecture_10/text_mining/data/data_ce_speech_article.rds")

dim(d) # Get dimension of dataset d
names(d) # Get column names of dataset d

# str(d) # chr 
# DT::datatable(d |> slice(1:5))
```

## Getting to know strings

-   Strings, when referred to in R, should be moderated by quotation marks (single or double are equivalent).

-   Everything is the same when you moderate them by single quotation marks.

-   String variables in a column are marked by variable types `chr` or `character`

## Case 1: When are the speeches/articles delivered?

Take a look at at the `date_of_speech` variable:

```{r}
d |> select(date_of_speech)
```

The `date_of_speech` variable is currently treated as a string and we are unable to get valuable information from it. Our first task is to use R's string operation to subtract inforamation of interest from this string-type indicator of date.

### Subtract strings using locations with `str_sub`

With start with the simplest string operation: Subtracting information of interest from a string based on locations.

```{r}
# Case: Getting YEAR
d |>
  select(date_of_speech) |>
  # create a new variable called YEAR, that gets the  the last but three to the last characters of the string
  mutate(YEAR = str_sub(date_of_speech, start = -4, end = -1))

# Case: Getting DAY
d |>
  select(date_of_speech) |>
  # DAY (of month): the first two characters of the strings
  mutate(DAY = str_sub(date_of_speech, start = 1, end = 2))
```

Note:

-   The key is the position

    -   `start = -4` means starting from the last but 4-1=3 position.

    -   `end = -1` means ending at the last position

-   If you take away the negative sign, then you are telling R to start/end at certain positions of the string. This will be useful for our extraction of the "day of month" information

### Remove and replace

Up next, we discuss methods to remove certain patterins from a string or replace a pattern with another pattern. Specifically, we will discuss the following functions:

-   `str_remove` and `str_remove_all`

-   `str_replace` and `str_replace_all`

-   If you use `str_remove` instead of `str_remove_all`, it will only remove the first occurance

```{r, echo=FALSE}
# remove "." from month
d_date |>
  mutate(MONTH_t = str_remove_all(MONTH, "\\."))
```

```{r}
# to get dates
d_date = d |>
  select(filename, date_of_speech) |>
  mutate(
    YEAR = str_sub(date_of_speech, start = -4, end = -1),
    MONTH = str_sub(date_of_speech, start = 3, end = -5),
    DAY = str_sub(date_of_speech, start = 1, end = 2)
    )
```

#### Exercise: Complete the cleaned data

Requirement:

-   Name the processed dataset `d_date_1`

-   Clean out the `.`'s

-   Convert `YEAR` `MONTH` `DAY` to numeric variables (hint: `as.numeric()`)

-   Optional: Provide summary statistics for `YEAR` `MONTH` `DAY` respectively

```{r}
# create processed data set d_date_1
d_date_1 <- d_date |>
  
  #clean out the "."s and convert variables to numeric variables
  mutate_at(vars(YEAR, MONTH, DAY), ~as.numeric(str_remove_all(., "\\.")))
```

```{r}
print("year summary")
summary(d_date_1$YEAR)

print("month summary")
summary(d_date_1$MONTH)

print("day summary")
summary(d_date_1$DAY)
```

### Replace patterns in strings

#### Case: Replace `.` by `-`

For demonstration purpose, let's do something meaningless first: Replace `.` by `-` for the `MONTH` variable.

```{r}
# first occurance
d_date |>
  mutate(MONTH = str_replace(MONTH, "\\.", "-"))
```

```{r}
# replace all
d_date |>
  mutate(MONTH = str_replace_all(MONTH, "\\.", "-"))
```

### Regular Expression

See Second page of <https://github.com/rstudio/cheatsheets/blob/main/strings.pdf>

Regular expression help you specify patterns of interest more effectively in strings.

-   How can we specify "ending `.` in `MONTH`?" Check the cheat sheet.

-   How can we specify "leading `.` in `MONTH`?" Check the cheat sheet.

```{r}
# Example: Removing the ending .
d_date |>
  mutate(MONTH_t = str_remove(MONTH, "\\.$"))

# Example: Removing the starting .
d_date |>
  mutate(MONTH_t = str_remove(MONTH, "^\\."))
```

A trick to see you have have found the right pattern in the string: `str_view`

```{r}
# Check the matched pattern in the first three entries of MONTH
# As an aside: "$" following dataframe's name select the variable; [1:3] selects the first three elements
str_view_all(d_date$MONTH[1:3], "\\.$", html = TRUE)

# This function will be more handy when your text is more complicated.
str_view_all(d$title[1:3], "Article by CE:", html = TRUE)
```

**Exercise**

1.  Replace the leading `.` of `MONTH` by 0
2.  Clean the `DAY` variable in the same way

```{r}
d_date |>
  mutate(
    MONTH = str_replace(MONTH, "^\\.", "0"),
    DAY = str_replace(DAY, "^\\.", "0")
  )
```

### Extract information from strings

What if we want to extract the numbers from dates? Use `str_extract`

```{r}
# Extract the first element found
d_date |>
  select(date_of_speech) |>
  mutate(date_of_speech_extract = 
           str_extract(date_of_speech, "[0-9]+"))

# Extract all the elements
d_date |>
  select(date_of_speech) |>
  mutate(date_of_speech_extract = str_extract_all(date_of_speech, "[0-9]+"))

```

Frequently seen output of string operation: A list column. We need to "unnest" list columns.

-   `unnest_wider` is typically used to unnest columns that have the same number of elements.

-   `unnest_longer` is typically used to unnest outputs that have various number of elements.

```{r}
# unnest_longer
# When unsure about the number of elements extracted. 
d_date |>
  select(date_of_speech) |>
  mutate(date_of_speech_extract = str_extract_all(date_of_speech, "[0-9]+")) |>
  unnest_longer(date_of_speech_extract)


# unnest_wider
d_date |>
  select(date_of_speech) |>
  mutate(date_of_speech_extract = str_extract_all(date_of_speech, "[0-9]+")) |>
  unnest_wider(date_of_speech_extract, names_sep = "_")

```

### Merge and Split Strings

Using the `str_` functions

-   Function that splits your strings using certain "separator": `str_split`

-   Function that merges your strings: `str_c`

```{r}
# Split the date variable using the separator
# Spearating the string using "." as spearators
d_date_result = d_date |>
  select(date_of_speech) |>
  mutate(
    date_of_speech_ex = str_split(date_of_speech, "\\.")
  ) |>
  unnest_wider(date_of_speech_ex, names_sep = "_")

a = 2
# evquivalent to
a <- 2
# equivalent to 
2 -> a
# Not equilalent to
# 2 = a # Error

```

```{r}
# Practice string merging with str_c
d_date_result

# Put together YEAR-MONTH-DAY

?str_c # paste, paste0

d_date_result |>
  mutate(
    date_merge = str_c(date_of_speech_ex_3, 
                       "-",
                       date_of_speech_ex_2, 
                       "-",
                       date_of_speech_ex_1
                       )
  )

d_date_result |>
  mutate(
    date_merge = str_c(date_of_speech_ex_3, 
                       "-",
                       str_pad(date_of_speech_ex_2, "0"), 
                       "-",
                       str_pad(date_of_speech_ex_1, "0")
                       )
  )
```

## Case 2: Wrangling the Title

Exercise:

-   Separate speeches and articles

-   Get speeches' locations

-   Identify policy addresses

-   Identify COVID-related speeches and article

```{r}
# Separate articles and speech

# Generate an variable indicating whether a piece is an article or a speech
d_2 = d |>
  mutate(
    article = str_extract(title, "Article by CE"),
    speech = str_extract(title, "Speech by CE")
  )

# Filter only articles
d |>
  filter(str_detect(title, "Article by CE"))

```

```{r}

```

## Case 3: Wrangling the Main Text

Exercises:

-   Remove redundant spaces

-   Remove line breaks and tabulation (`\n` `\r` `\t`)

-   Get dates of speech

-   Remove procedural contents from speeches

-   Get all numbers mentioned

-   Get all money mentioned

```{r}

```
