---
title: "Lecture 6 Notes Data Wrangling"
author: "Wong Nicole"
output: html_document
date: "2023-10-12"
---

# Data Wrangling

## Importing Data

```{r}
#import VDem data
library(tidyverse)
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
d |> print(n = 3)

#create new data set focusing on economic indicators

d_gdp <- d |> 
  select(country_text_id, year, e_gdp, e_gdppc) |>
  rename("gdp" = "e_gdp", "gdppc" = "e_gdppc")

d_gdp |> print(n = 3)
```

## Reshaping Data

### Why do we reshape data?

1.  Data cleaning
2.  Data visualization (e.g. some functions only take tables shaped in a specific way)
3.  Data sharing (e.g. you want to export the data for human readers)

### Examples

#### Wide to Long: `pivot_longer`

Example: one row for each variable

rows: country-gdp & country-gdppc

```{r}
d_gdp_long <- d_gdp |>
  pivot_longer(cols = c("gdp", "gdppc"), 
               names_to = "variable", values_to = "value")

d_gdp_long |> print(n = 4)
```

#### Long to Wide: `pivot_wider`

```{r}
#reverse the d_gdp_long
d_gdp_wide_1 <- d_gdp_long |>
  pivot_wider(names_from = "variable", values_from = "value")

d_gdp_wide_1 |> print(n = 4)
```

#### Make each unique value in a column a column

Example: Make year the column variable, i.e.

-   rows: country-gdp & country-gdppc

-   columns: year

```{r}
#make year the column
d_gdp_wide_2 <- d_gdp_long |>
  pivot_wider(names_from = "year", values_from = "value")

d_gdp_wide_2 |> print(n = 2)
```

Example: Make the country the column variable

```{r}
d_gdp_wide_3 <- d_gdp_long |>
  pivot_wider(names_from = "country_text_id", values_from = "value")

d_gdp_wide_3 |> print(n = 2)
```

## Stack Tables

-   Merge (or "stack") two data sets

-   \*\*\*Only when you know you can just safely stack them together as this is quite error prone!

### Task

-   Let's say we want to merge your GDP data `d_gdp` with some additional datasets that *you know* you can just safely stack together.

-   Example

    -   Merge with GDP data from 1906 to 1983

    -   Merge with education and Freedom House data from 1984 to 2022

#### Load New Data

-   Aim: Stack data vertically

-   GDP data from

    -   1945-1983

    -   1906-1944

```{r, message=FALSE}
d_gdp_1945 <- 
  read_csv("_DataPublic_/vdem/1945_1983/vdem_1945_1983_external.csv") |>
  select(country_text_id, year, e_gdp, e_gdppc) |>
  rename("gdp" = "e_gdp", "gdppc" = "e_gdppc")

d_gdp_1906 <- 
  read_csv("_DataPublic_/vdem/1906_1944/vdem_1906_1944_external.csv") |>
  select(country_text_id, year, e_gdp, e_gdppc) |>
  rename("gdp" = "e_gdp", "gdppc" = "e_gdppc")

d_gdp_1945 |> print(n = 2) 

d_edu <- d |>
  select(e_peaveduc, e_peedgini) |>
  rename("edu_15" = "e_peaveduc", "edu_gini" = "e_peedgini")

d_fh <- d |>
  select(starts_with("e_fh")) |>
  rename("fh_CivilLiberty" = "e_fh_cl", "fh_PoliticalRight" = "e_fh_pr",
         "fh_RuleOfLaw" = "e_fh_rol", "fh_Status" = "e_fh_status")

d_fh |> print(n = 2)
```

#### `bind_rows`

-   Match columns with the same names

-   Includes columns from BOTH data sets, set missing values as NA

```{r}
d_gdp_1945_2022 <- bind_rows(d_gdp, d_gdp_1945) # can take multiple data frames
d_gdp_1945_2022 |> print(n = 3)
unique(d_gdp_1945_2022$year) |> sort()

d_gdp_1906_2022 <- bind_rows(d_gdp, d_gdp_1945, d_gdp_1906) # can take multiple data frames
d_gdp_1906_2022 |> print(n = 3)
unique(d_gdp_1906_2022$year) |> sort()
```

#### `bind_cols`

```{r}
d_gdp_edu_fh <- bind_cols(d_gdp, d_edu, d_fh) # can take multiple data frames
d_gdp_edu_fh |> print(n = 3)
names(d_gdp_edu_fh)
```

## Join Tables

### Tasks: Understand the behavior of different `join_` functions

-   `left_join`: Merge and only keep observations that appear in the left-hand-side table.

-   `right_join`: Merge and only keep observations that appear in the right-hand-side table.

-   `inner_join`: Merge and only keep observations that appear in both tables.

-   `full_join`: Merge and keep observations that appear either table.

-   `anti_join`: Filter out observations that does appears in the right-hand-side table

-   `semi_join`: Filter out observations that does not appear in the right-hand-side table

### Task 1: The Case

Join two datasets from the V-Dem data using the above different `join_` functions

-   *GDP* data from 2000-2022

-   *GDP per capita* data from 1984 to 2010

```{r}
#load data set
d_gdp_2000_2022 <- d |> filter(year %in% 2000:2022) |>
  select(country_text_id, year, e_gdp) |> rename("gdp" = "e_gdp")

d_gdppc_1984_2010 <- d |> filter(year %in% 1984:2010) |>
  select(country_text_id, year, e_gdppc) |> rename("gdppc" = "e_gdppc")

d_gdp_2000_2022 |> print(n = 2)
d_gdppc_1984_2010 |> print(n = 2)
```

#### `left_join/ right_join`

-   merge and only keep identifiers that show up on the left/right table

```{r}
#left join
d_lj <- d_gdp_2000_2022 |> 
  left_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_lj |> print(n = 2)
unique(d_lj$year) |> sort()
```

```{r}
#right join
d_rj <- d_gdp_2000_2022 |> 
  right_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_rj |> print(n = 2)
unique(d_rj$year) |> sort()
```

### `inner_join`

-   merge and only keep identifiers that show up in [**both**]{.underline} tables

```{r}
d_ij <- d_gdp_2000_2022 |> 
  inner_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_ij |> print(n = 2)
unique(d_ij$year) |> sort()
```

### `full_join`

-   merge and only keep all identifiers

```{r}
d_fj <- d_gdp_2000_2022 |> 
  full_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_fj |> print(n = 2)
unique(d_fj$year) |> sort()
```

### `semi_join`

-   filter out observations that appears in the right table

```{r}
d_sj <- d_gdp_2000_2022 |> 
  semi_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_sj |> print(n = 2)
unique(d_sj$year) |> sort()
```

### `anti_join`

-   filter out observations that does [**NOT**]{.underline} appears in the right table

```{r}
d_aj <- d_gdp_2000_2022 |> 
  anti_join(d_gdppc_1984_2010, by = c("country_text_id", "year"))

d_aj |> print(n = 2)
unique(d_aj$year) |> sort()
```

### Join by Identifiers with Different Variable Names

If the identifiers have different names, you have two options:

(1) Rename it beforehand

```{r}
#create artificial identifier
d_gdppc_1984_2010_t <- d_gdppc_1984_2010 |> 
  rename("country_id" = "country_text_id")

# Option 1: Rename the variables
d_aj_t <- d_gdp_2000_2022 |> 
  rename("country_id" = "country_text_id") |>
  anti_join(d_gdppc_1984_2010_t, by = c("country_id", "year"))
```

(2) specify the `by =` argument differently

```{r}
# Option 2: Specify the "by =" argument with a *named vector*
d_aj_t_2 <- d_gdp_2000_2022 |> 
  anti_join(d_gdppc_1984_2010_t, 
            by = c("country_text_id" = "country_id", 
                   "year" = "year"))
```

### Many-to-One Join: Repeat!

Calculate each country's average 1984-2010 *GDP per capita* and merge it with our annual GDP data from 2000 to 2022.

```{r}
d_gdppc_1984_2010_avg <- d_gdppc_1984_2010 |> group_by(country_text_id) |>
  summarise(gdppc_1984to2010 = mean(gdppc, na.rm = TRUE))
d_gdppc_1984_2010_avg |> print(n = 2)
```

```{r}
d_lj_ManyToOne <- d_gdp_2000_2022 |> 
  left_join(d_gdppc_1984_2010_avg, by = "country_text_id")
d_lj_ManyToOne |> print(n = 2)
```

### Remarks about Join

-   Joining tables may be error-prone to avoid replication issue

-   Good habits

    -   Add suffixes/ prefixes indicating data sources

    -   Add binary indicators (1/0) indicating from in which data set each observation is available

#### Adding Binary Indicators

```{r}
# The d_gdp_2000_2022 data are from V-Dem
d_gdp_2000_2022_t <- d_gdp_2000_2022 |> mutate(source_vdem = 1)

# *Pretend* that the d_gdppc_1984_2010 data are from the World Bank
d_gdppc_1984_2010_t <- d_gdppc_1984_2010 |> mutate(source_wb = 1)

d_fj_habit <- d_gdp_2000_2022_t |> 
  full_join(d_gdppc_1984_2010_t, by = c("country_text_id", "year"))

d_fj_habit |> print(n = 3)
```

```{r}
#Identify data overlapping
d_fj_habit |>
  group_by(source_vdem, source_wb) |>
  count()
```

#### Adding Suffixes/Prefixes

```{r}
d_gdp_2000_2022_rn <- d_gdp_2000_2022 |> 
  rename("vdem_gdp" = "gdp")
  # rename_at(vars(-c("country_text_id", "year")), ~str_c("vdem_", .))

d_gdppc_1984_2010_rn <- d_gdppc_1984_2010 |> 
  rename("wb_gdppc" = "gdppc")
  # rename_at(vars(-c("country_text_id", "year")), ~str_c("wb_", .))

d_fj_habit_2 <- d_gdp_2000_2022_rn |>
  full_join(d_gdppc_1984_2010_rn, by = c("country_text_id", "year"))

d_fj_habit_2 |> print(n = 3)
```

# Save Outputs

## Saving Your Outputs after Data Wrangling

You can save your clean data in a variety of formats. I will highlight two most popular options

-   `.csv` "comma-separated values," readable by Excel or a text editor

-   `.rds` "R data serialization," readable by R only

```{r}
# Save to a .csv file
write_csv(d_gdp_1945_2022, "Lec_06/2_data_wrangling_3/data/gdp_1945_2002.csv")

# Save to a .rds file
saveRDS(d_gdp_1945_2022, "Lec_06/2_data_wrangling_3/data/gdp_1945_2002.rds")
```

## Saving Your Outputs after Data Wrangling

You can re-load saved `.csv` and `.rds` files using `read_csv` and `readRDS` respectively

```{r}
# Read a .csv file
d_read_1 <- read_csv("Lec_06/2_data_wrangling_3/data/gdp_1945_2002.csv")

# Read a .rds file
d_read_2 <- readRDS("Lec_06/2_data_wrangling_3/data/gdp_1945_2002.rds")
```

## Comparison: .csv vs .rds

+--------+----------------------------------------------------------+------------------------------------------------------------------------------------------+---------------------------------------------------------+
| Type   | Pro                                                      | Con                                                                                      | When?                                                   |
+========+==========================================================+==========================================================================================+=========================================================+
| `.csv` | -   Readable outside R                                   | -   Variable types may change when you read it back if you do not carefully specify them | -   Simple data types                                   |
|        |                                                          |                                                                                          |                                                         |
|        | -   Conveniently convertible to Excel files              | -   Error-prone with *text* data (encoding, line breaks etc.)                            | -   Want to manually examine it outside R (e.g., Excel) |
|        |                                                          |                                                                                          |                                                         |
|        |                                                          | -   (Maybe) takes longer to read                                                         | -   Want to share it with non-R users                   |
+--------+----------------------------------------------------------+------------------------------------------------------------------------------------------+---------------------------------------------------------+
| `.rds` | -   **Replicable**: Get precisely how the data are saved | -   Can't read `.rds` outside R                                                          | -   Complex combination of data types                   |
|        |                                                          |                                                                                          |                                                         |
|        | -   Smaller files (if stick with default compression)    |                                                                                          | -   Simply saving for your future use in R              |
|        |                                                          |                                                                                          |                                                         |
|        | -   (Sometimes) faster read/write                        |                                                                                          | -   Large dataset and you want to save space            |
|        |                                                          |                                                                                          |                                                         |
|        |                                                          |                                                                                          | -   Text data                                           |
|        |                                                          |                                                                                          |                                                         |
|        |                                                          |                                                                                          | -   **Always safer**                                    |
+--------+----------------------------------------------------------+------------------------------------------------------------------------------------------+---------------------------------------------------------+
