---
title: "Data Wrangling 2: Reshape and Combine Tables"
author: "Haohan Chen"
date: "Last update: `r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document: default
  pdf_document: default
  md_document: default
knit: (function(inputFile, encoding){rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = getwd())})
editor_options: 
  chunk_output_type: console
---

## Objective

This lecture continues our introduction to data wrangling with R. Using the *V-Dem* data as an example, we will learn how to *reshape* and *merge* datasets using a set of [`tidyverse`](https://www.tidyverse.org/) functionality. Specifically, we will focus on functions...

1.  ... to *reshape* a table (long \<-\> wide) with `pivot_longer` and `pivot_wider`
2.  ... to *stack* tables by row or by column with `bind_rows` and `bind_cols` (or, alternatively, `cbind` and `rbind`)
3.  ... to *merge* two tables with `inner_join`, `full_join`, `left_join`, `right_join`, `semi_join`, and `anti_join`

## Further Reading

-   R for Data Science (2e) Chapters 6 and 20: <https://r4ds.hadley.nz/>

-   `dplyr` cheatsheet (the "Combine Tables" section on p. 2)

-   `tidyr` cheatsheet (the "Reshape Data" section on p. 1): <https://rstudio.github.io/cheatsheets/html/tidyr.html>

## In-Class Demo

```{r}
library(tidyverse)
```

```{r}
d <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
```

### 1. Reshape the V-Dem dataset

```{r}
#want each row contain country-year-variable
d |> select(country_name) |> distinct()

#want each row contain country-year-variable where country name starts with E
d_subset <- d |> select(country_name, year, starts_with("e_"))
#starts_with e refers to a list of COLUMNS starting with e

d_subset_long <-
  d_subset |> pivot_longer(col = starts_with("e_"))
```

```{r}
d_subset_wide_year <-
  d_subset_long |> pivot_wider(names_from = year, values_from = value)
```

### 2. Stack multiple subsets of the V-Dem datasets

```{r}
d_VdemIndex <- read_csv("_DataPublic_/vdem/1984_2022/vdem_1984_2022_external.csv")
bind_cols(d, d_VdemIndex)
#R automatically merges variables that have the same name
```

```{r}
#want to stack two tables by row
d_1945_1983 <- read_csv("_DataPublic_/vdem/1945_1983/vdem_1945_1983_external.csv", show_col_types = FALSE)

bind_rows(d, d_1945_1983)
```

### 3. Join multiple regional subsets of the V-Dem datasets

```{r}
# INSERT CODE
```

### 4. Validate the GDP data in V-Dem with World Bank data

**Task:** There are many different "versions" of GDP data. I wonder whether the GDP data in the V-Dem dataset is reliable. So I would like to validate it with [data from the World Bank](https://data.worldbank.org/).

#### Download World Bank Data

We will start the adventure by downloading World Bank data.

```{r}
# Install the WDI package that helps fetch data from the World Bank dataset
# See: https://github.com/vincentarelbundock/WDI

# install.packages("WDI")

# Note: Comment out the above "install.packages" command after you are done with installing the package.
```

```{r}
library(WDI)
```

```{r}
# Search for GDP related data
wb_gdpdata_list <- WDIsearch("gdp")

str(wb_gdpdata_list)

# Narrow down to indicators of GDP (I have done some pre-screening)
wb_gdpdata_list_s <- wb_gdpdata_list |> filter(str_detect(indicator, "^NY\\.GDP"))
```

```{r}
# Download GDP-related data
wb_gdpdata <- WDI(
  indicator = c("NY.GDP.MKTP.PP.KD", "NY.GDP.PCAP.PP.KD"), 
  country = "all", 
  start = 1984, end = 2022)
```

```{r}
# Remove the intermediate data we no longer need.
rm(wb_gdpdata_list, wb_gdpdata_list_s)
```

To match two datasets from two different sources, we should always check whether the "identifiers" are consistent. In our case, are names of countries specified in the same way in the V-Dem and the World Bank dataset?

```{r}
# Check the specification of country names.

```

#### Find Country Identifiers

When it comes to matching countries, country codes are usually more reliable. The problem is that we do not have country codes in the V-Dem data. An R package named `countrycode` can help.

```{r}
# install.packages("countrycode")
# See how you may use the package: https://github.com/vincentarelbundock/countrycode
```

```{r}
# INSERT CODE: Use countrycode to make country code indicators
```

#### Join and Compare

Now that we have cleaned the World Bank data, our final step is to join it with the World Bank data and compare the GDP and GDP per capita indicators from the two sources.

```{r}
# INSERT CODE: Join the two datasets
```

```{r}
# INSERT CODE: Compare the two datasets
```
