---
title: "Lecture 8 Notes (One category)"
author: "Wong Nicole"
date: "2023-11-02"
output: html_document
---

# Load Data

```{r}
library(tidyverse)
theme_set(theme_bw()) # Set my default theme for the whole document

d0 <- readRDS("Lecture_8/data/wealth_and_health.rds") # check that it exists in data file
#d |> print(n = 3)
```

```{r}
# quick summary
summary(d0$region)
```

```{r}
d <- d0 |>
  mutate(
    region = case_match(
      region,
      1 ~ "Western Europe", 2 ~ "Northern Europe", 3 ~ "Southern Europe",
      4 ~ "Eastern Europe", 5 ~ "Northern Africa", 6 ~ "Western Africa",
      7 ~ "Middle Africa", 8 ~ "Eastern Africa", 9 ~ "Southern Africa",
      10 ~ "Western Asia", 11 ~ "Central Asia", 12 ~ "Eastern Asia",
      13 ~ "South-Eastern Asia", 14 ~ "Southern Asia", 15 ~ "Oceania",
      16 ~ "North America", 17 ~ "Central America", 18 ~ "South America", 19 ~ "Caribbean",
      .default = NA))
```

# 1 Categorical Variable

## Summary

```{r}
# no. of observations
length(d$region)
```

```{r}
# no. of missing variables
is.na(d$region) |> sum()
```

```{r}
table(d$region, useNA = "always")
```

```{r}
table(d$region, useNA = "always") |> prop.table()
```

## Bar Charts

```{r}
# vertical plots
# use theme to adjust the labels (prevent overlapping)
d |> ggplot(aes(x = region)) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

```{r}
# horizontal plots
d |> ggplot(aes(x = region)) + geom_bar() +
  coord_flip() +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Ordered Bar Charts

Order by frequency

-   Use `fct_infreq` for descending order

-   Use `fct_rev` for ascending order

```{r}
# ordered bar charts
d |> ggplot(aes(x = fct_infreq(region))) + geom_bar() + #desc. order
  #ggplot(aes(x = fct_rev(fct_infreq(region)))) + geom_bar() + #asc. order
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

Custom orders (consistency)

```{r}
#define fixed orders
region_levels <- c(
  "Western Europe", "Northern Europe", "Southern Europe", "Eastern Europe",
  "Northern Africa", "Western Africa", "Middle Africa", "Eastern Africa", "Southern Africa",
  "Western Asia", "Central Asia", "Eastern Asia", "South-Eastern Asia", "Southern Asia",
  "Oceania", 
  "North America", "Central America", "South America", "Caribbean")

d |>
  mutate(region = factor(region, levels = region_levels)) |>
  ggplot(aes(x = region)) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Calculate frequency \*

In all the previous examples, we let `ggplot` calculate the frequency for us. To allow for more customization, a better way is to calculate the frequencies manually before using `geom_bar`.

```{r}
d |> group_by(region) |> summarise(n_obs = n())
```

```{r}

# ordered by n_obs (descending order)

d |> group_by(region) |> summarise(n_obs = n()) |>
  ggplot(aes(x = reorder(region, -n_obs), y = n_obs)) + geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")  
```

## Re-categorize

Sometimes there may be too many categories. You can re-categorize (e.g. from regions to continents, labelled as region_higher)

```{r}
d <- d |>
  mutate(region_higher = case_match(
    region,
    c("Western Europe", "Northern Europe", "Southern Europe", "Eastern Europe") ~ "Europe",
    c("Northern Africa", "Western Africa", "Middle Africa", "Eastern Africa", "Southern Africa") ~ "Africa",
    c("Western Asia", "Central Asia", "Eastern Asia", "South-Eastern Asia", "Southern Asia") ~ "Asia",
    c("Central America", "South America", "Caribbean") ~ "Latin America",
    .default = region), .before = region)

table(d$region_higher)
```

```{r}
d |> group_by(region_higher) |> summarise(n_obs = n()) |>
  ggplot(aes(y = reorder(region_higher, n_obs), x = n_obs)) + geom_bar(stat = "identity") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region")  
```

## Adding a Second Variable

### Stacked

We can use a stacked bar chart

-   use `fill = factor(variable x)` as demoncracy_binary is a binary variable

```{r}
table(d$region_higher)
```

```{r}
# stacked bar chart
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = factor(democracy_binary))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")  
```

Alternatively, we can change 0, 1, NA to meaningful names

-   Mutate to change binary from 1/0/ to Yes/No

```{r}
# stacked bar chart with renamed binary
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")
```

### Stacked + poportion

100% stacked bar charts can be used to visualize proportion

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  group_by(region_higher) |> mutate(prop_obs = n_obs / sum(n_obs) * 100) |> # Calculate proportion
  ggplot(aes(x = prop_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Percentage (%)", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")

```

```{r}
# with annotation
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  group_by(region_higher) |> mutate(prop_obs = n_obs / sum(n_obs) * 100) |>
  ggplot(aes(x = prop_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n_obs), position = position_stack(vjust = 0.5), size = 2) +
  labs(y = "Region", x = "Proportion", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")

```

# 1 cat. + 1 quant.

## Box Plots

-   use `coord_flip()` to flip the chart

```{r}
# plot verticle box graphs
d |> ggplot(aes(x = region_higher, y = life_expectancy)) + 
  geom_boxplot() + coord_flip() +
  labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

```{r}
# ordered by median
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_boxplot() + coord_flip() +
  labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

-   use `geom_jitter()` to add data points

```{r}
# show data points
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA. Essential for the reordering to work!
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_boxplot() + geom_jitter(position = position_jitter(0.3), alpha = 0.1) + # geom_jitter() does the trick.
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Violin Plot

```{r}
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_violin() +
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

```{r}
# combine violin plot with box plot
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_violin() + geom_boxplot(width = 0.15, fill = "gray") + # Note: Place geom_boxplot after geom_violin.
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

# 1 cat. + 2 quant.
