---
title: "Lecture 8 Notes (One category)"
author: "Wong Nicole"
date: "2023-11-02"
output: html_document
---

# Load Data

```{r}
library(tidyverse)
theme_set(theme_bw()) # Set my default theme for the whole document

d0 <- readRDS("Lecture_8/data/wealth_and_health.rds") # check that it exists in data file
#d |> print(n = 3)
```

```{r}
# quick summary
summary(d0$region)
```

```{r}
d <- d0 |>
  mutate(
    region = case_match(
      region,
      1 ~ "Western Europe", 2 ~ "Northern Europe", 3 ~ "Southern Europe",
      4 ~ "Eastern Europe", 5 ~ "Northern Africa", 6 ~ "Western Africa",
      7 ~ "Middle Africa", 8 ~ "Eastern Africa", 9 ~ "Southern Africa",
      10 ~ "Western Asia", 11 ~ "Central Asia", 12 ~ "Eastern Asia",
      13 ~ "South-Eastern Asia", 14 ~ "Southern Asia", 15 ~ "Oceania",
      16 ~ "North America", 17 ~ "Central America", 18 ~ "South America", 19 ~ "Caribbean",
      .default = NA))
```

# 1 Categorical Variable

## Summary

```{r}
# no. of observations
length(d$region)
```

```{r}
# no. of missing variables
is.na(d$region) |> sum()
```

```{r}
table(d$region, useNA = "always")
```

```{r}
table(d$region, useNA = "always") |> prop.table()
```

## Bar Charts

```{r}
# vertical plots
# use theme to adjust the labels (prevent overlapping)
d |> ggplot(aes(x = region)) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

```{r}
# horizontal plots
d |> ggplot(aes(x = region)) + geom_bar() +
  coord_flip() +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Ordered Bar Charts

Order by frequency

-   Use `fct_infreq` for descending order

-   Use `fct_rev` for ascending order

```{r}
# ordered bar charts
d |> ggplot(aes(x = fct_infreq(region))) + geom_bar() + #desc. order
  #ggplot(aes(x = fct_rev(fct_infreq(region)))) + geom_bar() + #asc. order
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

Custom orders (consistency)

```{r}
#define fixed orders
region_levels <- c(
  "Western Europe", "Northern Europe", "Southern Europe", "Eastern Europe",
  "Northern Africa", "Western Africa", "Middle Africa", "Eastern Africa", "Southern Africa",
  "Western Asia", "Central Asia", "Eastern Asia", "South-Eastern Asia", "Southern Asia",
  "Oceania", 
  "North America", "Central America", "South America", "Caribbean")

d |>
  mutate(region = factor(region, levels = region_levels)) |>
  ggplot(aes(x = region)) + geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")
```

## Calculate frequency \*

In all the previous examples, we let `ggplot` calculate the frequency for us. To allow for more customization, a better way is to calculate the frequencies manually before using `geom_bar`.

```{r}
d |> group_by(region) |> summarise(n_obs = n())
```

```{r}

# ordered by n_obs (descending order)

d |> group_by(region) |> summarise(n_obs = n()) |>
  ggplot(aes(x = reorder(region, -n_obs), y = n_obs)) + geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0)) +
  labs(x = "Region", y = "Count", title = "N. Country-Year by Region")  
```

## Re-categorize

Sometimes there may be too many categories. You can re-categorize (e.g. from regions to continents, labelled as region_higher)

```{r}
d <- d |>
  mutate(region_higher = case_match(
    region,
    c("Western Europe", "Northern Europe", "Southern Europe", "Eastern Europe") ~ "Europe",
    c("Northern Africa", "Western Africa", "Middle Africa", "Eastern Africa", "Southern Africa") ~ "Africa",
    c("Western Asia", "Central Asia", "Eastern Asia", "South-Eastern Asia", "Southern Asia") ~ "Asia",
    c("Central America", "South America", "Caribbean") ~ "Latin America",
    .default = region), .before = region)

table(d$region_higher)
```

```{r}
d |> group_by(region_higher) |> summarise(n_obs = n()) |>
  ggplot(aes(y = reorder(region_higher, n_obs), x = n_obs)) + geom_bar(stat = "identity") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region")  
```

## Adding a Second Variable

### Stacked

We can use a stacked bar chart

-   use `fill = factor(variable x)` as demoncracy_binary is a binary variable

```{r}
table(d$region_higher)
```

```{r}
# stacked bar chart
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = factor(democracy_binary))) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")  
```

Alternatively, we can change 0, 1, NA to meaningful names

-   Mutate to change binary from 1/0/ to Yes/No

```{r}
# stacked bar chart with renamed binary
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  ggplot(aes(x = n_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Count", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")
```

### Stacked + poportion

100% stacked bar charts can be used to visualize proportion (of democracy in each geographic region)

```{r}
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  group_by(region_higher) |> mutate(prop_obs = n_obs / sum(n_obs) * 100) |> # Calculate proportion
  ggplot(aes(x = prop_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(y = "Region", x = "Percentage (%)", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")

```

```{r}
# with annotation
d |> group_by(region_higher, democracy_binary) |> summarise(n_obs = n()) |>
  mutate(democracy_binary = case_match(democracy_binary, 1 ~ "Yes", 0 ~ "No", NA ~ NA)) |>
  group_by(region_higher) |> mutate(prop_obs = n_obs / sum(n_obs) * 100) |>
  ggplot(aes(x = prop_obs, y = region_higher, fill = democracy_binary)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n_obs), position = position_stack(vjust = 0.5), size = 2) +
  labs(y = "Region", x = "Proportion", title = "N. Country-Year by Region and Regime Type", fill = "Democracy")

```

# 1 cat. + 1 quant.

## Box Plots

-   use `coord_flip()` to flip the chart

```{r}
# plot verticle box graphs
d |> ggplot(aes(x = region_higher, y = life_expectancy)) + 
  geom_boxplot() + coord_flip() +
  labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

```{r}
# ordered by median
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_boxplot() + coord_flip() +
  labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

-   use `geom_jitter()` to add data points

```{r}
# show data points
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA. Essential for the reordering to work!
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_boxplot() + geom_jitter(position = position_jitter(0.3), alpha = 0.1) + # geom_jitter() does the trick.
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Violin Plot

```{r}
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_violin() +
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

```{r}
# combine violin plot with box plot
d |> 
  filter(!is.na(life_expectancy)) |> # Remove entries whose life_expectancy is NA
  ggplot(aes(x = fct_reorder(region_higher, life_expectancy, .fun=median, .desc = FALSE), y = life_expectancy)) + 
  geom_violin() + geom_boxplot(width = 0.15, fill = "grey") + # Note: Place geom_boxplot after geom_violin.
  coord_flip() + labs(title = "Life Expectancy by Region", x = "Region", y = "Life Expectancy")
```

## Colored Histogram

```{r}
d |>
  ggplot(aes(x = life_expectancy, fill = region_higher)) +
  geom_histogram(bins = 50) + 
  labs(title = "Life Expectancy by Region", fill = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

## Colored Frequency Polygon

```{r}
d |>
  ggplot(aes(x = life_expectancy, color = region_higher)) +
  geom_freqpoly(bins = 50) + 
  labs(title = "Life Expectancy by Region", fill = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

## Colored Density

```{r}
d |>
  ggplot(aes(x = life_expectancy, fill = region_higher)) +
  geom_density(alpha = 0.2) + 
  labs(title = "Life Expectancy by Region", fill = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

```{r}
# alternative design
d |>
  ggplot(aes(x = life_expectancy, fill = NULL, color = region_higher)) +
  geom_density() + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom")
```

## Facets

Does your figure look too "busy"? Separate them into different sub-figures!

```{r}
# columns or rows
d |>
  ggplot(aes(x = life_expectancy, fill = NULL)) +
  geom_density() + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom") +
  facet_grid(cols = vars(region_higher))
  #facet_grid(rows = vars(region_higher))
```

```{r}
# flexible
d |>
  ggplot(aes(x = life_expectancy, fill = NULL)) +
  geom_density() + 
  labs(title = "Life Expectancy by Region", color = "Region", x = "Life Expectancy") +
  theme(legend.position = "bottom") +
  facet_wrap(~region_higher, nrow = 2)
```

# 1 cat. + 2 quant.

## Scatter Plot

Change colour based on region

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(color = region_higher), alpha = 0.3, shape = 3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

Change shape based on region (may be difficult to view :) )

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

Or both

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher, color = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

Add trend lines

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher, color = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  geom_smooth(aes(color = region_higher)) +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)")
```

### Facets 🌈🌈🌈

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(shape = region_higher, color = region_higher), alpha = 0.3, size = 0.5, stroke = 1) +
  geom_smooth() +
  labs(x = "GDP per capita", y = "Life Expectancy", color = "Region", shape = "Region",
       title = "Wealth and Health in the World (1800-2019)") +
  facet_wrap(~region_higher, nrow = 2, scales = "free") +
  theme(legend.position = "none")
```

## Heatmap

```{r}
d |> filter(year >= 1984) |>
  group_by(region_higher, year) |> summarise(life_expectancy_avg = mean(life_expectancy, na.rm = TRUE)) |> mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = region_higher, fill = life_expectancy_avg)) + geom_tile() +
  theme(axis.text.x = element_text(angle = 45), legend.position = "bottom") +
  labs(x = "Year", y = "Region", fill = "Life Expectancy", title = "Life Expectancy by Region-Year")
```

```{r}
# colored heat map
d |> filter(year >= 1984) |>
  group_by(region_higher, year) |> summarise(life_expectancy_avg = mean(life_expectancy, na.rm = TRUE)) |> mutate(year = factor(year)) |>
  ggplot(aes(x = year, y = region_higher, fill = life_expectancy_avg)) + geom_tile() +
  theme(axis.text.x = element_text(angle = 45), legend.position = "bottom") +
  scale_fill_viridis_c(option = "A", direction = 1) +
  labs(x = "Year", y = "Region", fill = "Life Expectancy", title = "Life Expectancy by Region-Year")
```

## Facets: Density Plot Matrix `facet_grid`

```{r}
d |> filter(year %in% 2010:2020) |> ggplot(aes(x = life_expectancy)) + geom_density() +
  facet_grid(cols = vars(region_higher), rows = vars(year)) +
  labs(x = "Life Expectancy", y = "Density", title = "Life Expectancy by Region-Year (2010-2022)")
```

```{r}
# free scales on x
d |> filter(year %in% 2010:2020) |> ggplot(aes(x = life_expectancy)) + geom_density() +
  facet_grid(cols = vars(region_higher), rows = vars(year), scales = "free_x") +
  labs(x = "Life Expectancy", y = "Density", title = "Life Expectancy by Region-Year (2010-2022)")
```

```{r}
# free scales on y
d |> filter(year %in% 2010:2020) |> ggplot(aes(x = life_expectancy)) + geom_density() +
  facet_grid(cols = vars(region_higher), rows = vars(year), scales = "free_y") +
  labs(x = "Life Expectancy", y = "Density", title = "Life Expectancy by Region-Year (2010-2022)")
```
