---
title: "Lecture 7 Notes Data Visualisation with ggplot"
author: "Wong Nicole"
date: "2023-10-26"
output: html_document
---

## Load Data

```{r}
library(tidyverse)
theme_set(theme_bw()) # Set my default theme for the whole document
```

```{r}
d <- readRDS("Lecture_7/data/wealth_and_health.rds")
d |> print(n = 3)
```

## `ggplot` syntax

-   Consider: \"Painting\" your data on a canvas, layer by layer

-   Steps

    -   Specify the data you want to viaulize

    -   Specify the type(s) of visualization tools

    -   Label the figure (titles of the figure, axes, and legends)

-   Command: Use **`+`** to connect layers (ATTENTION: not `|>` )

```{r}
d |> # Specify the data frame to plot
  ggplot(aes(x = gdppc, y = life_expectancy)) + # Specify the variables to plot
  geom_point() + # Choose the visualization tool: point; if you do not include step -> empty
  labs(x = "GDP per capita", y = "Life Expectancy", # set up the axes' titles
       title = "Wealth and Health in the World (1800-2019)", # set up the figure's title
       caption = "By Haohan Chen. Data source: V-Dem v.13") # set up footnote
```

## Variable Types

Visualization tools to use largely depends on variable types

-   \"Quantitative\" (Continuous, Count, numeric)

    -   GDP per capita

    -   Life expectancy

    -   Population

    -   Infant mortality

-   Categorical

    -   Binary: Binary \"democracy\" indicator

    -   Nominal: Region

    -   Ordinal: Lexical Index of Electoral Democracy

### Quant x1

#### Summary Statistics

```{r}
summary(d$gdppc) # Quick summary
```

```{r}
length(d$gdppc) # Number of observations
```

```{r}
is.na(d$gdppc) |> sum() # Number of missing values
```

```{r}
mean(d$gdppc, na.rm = TRUE) # Mean
```

```{r}
var(d$gdppc, na.rm = TRUE) # Variance
```

```{r}
sd(d$gdppc, na.rm = TRUE) # Standard deviation
```

```{r}
quantile(d$gdppc, na.rm = TRUE) # Quantile
```

```{r}

quantile(d$gdppc, seq(0, 1, 0.1), na.rm = TRUE) # Quintile
```

#### Histogram

```{r}
d |> 
  ggplot(aes(x = gdppc)) +
  geom_histogram() + # Draw a histogram
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
# number of bins
# default = 30 bins (or bars)
d |> ggplot(aes(x = gdppc)) + geom_histogram(bins = 10) +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
# color
d |> ggplot(aes(x = gdppc)) + geom_histogram(bins = 50, fill = "blue", color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

### Density

```{r}
d |> 
  ggplot(aes(x = gdppc)) +
  geom_density() + # Draw a density plot
  labs(x = "GDP per capita", y = "Density", title = "Density of GDP per capita")
```

```{r}
# color to view area (shown in gray for e.g.)
d |> 
  ggplot(aes(x = gdppc)) +
  geom_density(fill = "gray", color = "red") + # Draw a density plot
  labs(x = "GDP per capita", y = "Density", title = "Density of GDP per capita")
```

### Skewed distribution

-   **Challenge:** Quantitative data may have skewed distributions, which can make the visualization uninformative.

-   **Solution:**

    -   Re-scale the axes

    -   Transform the variable(s)

#### Re-scale the axes

```{r}
d |> ggplot(aes(x = gdppc)) + geom_histogram() +
  scale_x_sqrt() + # Scale the x axis to make distribution easy to read
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
d |> ggplot(aes(x = gdppc)) + geom_histogram() +
  scale_x_log10() + # Scale the x axis to make distribution easy to read
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

#### Transform the variables

```{r}
d |> mutate(gdppc_sqrt = sqrt(gdppc)) |> # Transform the variable before visualization
  ggplot(aes(x = gdppc_sqrt)) + geom_histogram() +
  labs(x = "sqrt(GDP per capita)", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
d |> mutate(gdppc_log = log(gdppc)) |> # Transform the variable before visualization
  ggplot(aes(x = gdppc_log)) + geom_histogram() +
  labs(x = "log(GDP per capita)", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
#method only works with non negative-values
summary(d$gdppc)
```

### Annotating cases of interest

```{r}
d |> ggplot(aes(x = gdppc)) + geom_histogram() +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
d_anno <- d |> filter(country_text_id == "CHN", year == 2019) |> mutate(anno = paste(country_text_id, year), .after = year)
print(d_anno)
```

```{r}
# A "manual" approach
d |> ggplot(aes(x = gdppc)) + geom_histogram() + 
  geom_vline(aes(xintercept = 15.4), linetype = "dashed", color = "red") +
  annotate("text", x = 15.4, y = 7500, hjust = 0, label = "China 2019", color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
# An automated approach
d |> ggplot(aes(x = gdppc)) + geom_histogram() + 
  geom_vline(data = d_anno, aes(xintercept = gdppc), linetype = "dashed", color = "red") +
  geom_text(data = d_anno, aes(x = gdppc, y = 7500, label = anno), hjust = 0, color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

The autonated approach can be useful for making multiple annotations

```{r}
d_anno <- d |> filter(country_text_id %in% c("CHN", "JPN", "USA"), year == 2019) |> 
  mutate(anno = paste(country_text_id, year), .after = year)
print(d_anno)
```

```{r}
d |> ggplot(aes(x = gdppc)) + geom_histogram() + 
  geom_vline(data = d_anno, aes(xintercept = gdppc), linetype = "dashed", color = "red") +
  geom_text(data = d_anno, aes(x = gdppc, y = 7500, label = anno), hjust = 0, color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

```{r}
# also works for density graphs
d |> ggplot(aes(x = gdppc)) + 
  geom_density() + 
  geom_vline(data = d_anno, aes(xintercept = gdppc), linetype = "dashed", color = "red") +
  geom_text(data = d_anno, aes(x = gdppc, y = 0.15, label = anno), hjust = 0, color = "red") +
  labs(x = "GDP per capita", y = "Count", title = "Histogram of GDP per capita")
```

### Quan x2

### Scatter Plot

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, color = "navy", shape = 3, size = 0.5, stroke = 1) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")

```

```{r}
# trend line
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_smooth() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")

```

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_smooth(se = FALSE, color = "red") +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_smooth(method = "lm") +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

#### Skewed Distribution: Re-scale Axes

```{r}
# more detailed than all data points being clustered on the left side
# please add footnote that it is scaled to not mislead the audience
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, color = "navy", shape = 3, size = 0.5, stroke = 1) +
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

### Skewed Distribution: Transform variables

```{r}
d |> 
  mutate(gdppc_log = log(gdppc)) |>
  ggplot(aes(x = gdppc_log, y = life_expectancy)) +
  geom_point(alpha = 0.3, color = "navy", shape = 3, size = 0.5, stroke = 1) +
  labs(x = "log(GDP per capita)", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

### Annotate Data Points: `geom_text/geom_label`

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_text(data = d_anno, aes(x = gdppc, y = life_expectancy, label = anno), color = "red", size = 2.5) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

```{r}
# includes background --> more clear?
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  geom_label(data = d_anno, aes(x = gdppc, y = life_expectancy, label = anno), color = "red", size = 2.5) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")

```

\*geom_text may have strange outputs/overlaps for multiple data points

#### Annotate MANY Data Points: `geom_label_repel`

```{r}
library(ggrepel)
d_anno_2 <- d |> filter(year >= 2010, country_text_id == "CHN") |> 
  mutate(anno = paste(country_text_id, year), .after = year)
```

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(alpha = 0.3, shape = 3, size = 0.5) +
  # geom_point(data = d_anno_2, aes(x = gdppc, y = life_expectancy), color = "red", size = 0.5) +
  geom_label_repel(data = d_anno_2, aes(x = gdppc, y = life_expectancy, label = anno), 
                   color = "red", size = 2, max.overlaps = 20) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

## 2D "Histogram\"

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_bin_2d(bins = 80) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_bin_2d(bins = 80) + scale_x_log10() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

```{r}
# colour! :)
# scale_fill_viridis_c()
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_bin_2d(bins = 80) + scale_x_log10() +
  scale_fill_viridis_c() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```

```{r}
d |> ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_hex(bins = 80) + scale_x_log10() +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (1800-2019)")
```
