---
title: "Lecture 7 Notes Data Visualisation & Correlation"
author: "Wong Nicole"
date: "2023-10-26"
output: html_document
---

## Load Data

```{r}
library(tidyverse)
theme_set(theme_bw()) # Set my default theme for the whole document
d <- readRDS("Lecture_7/data/wealth_and_health.rds")
d |> print(n = 3)
```

## Summary Statistics: Pearson Correlation

Pearson correlation (aka. Pearson\'s *r*), ranging from -1 to 1, is our typical indicator of the correlation between two quantitative variables.

To use Pearson correlation, the two variables should meet the following assumptions:

-   Continuous

-   Linear relationship

-   No outlier

-   Normally distributed

```{r}
cor(d$life_expectancy, d$gdppc, use = "complete.obs", method = "pearson")
```

```{r}
cor.test(d$life_expectancy, d$gdppc, use = "complete.obs", method = "pearson")
```

## Summary Statistics: Spearman Correlation

If the data seem to break at least one of the assumptions for Pearson correlation, you may use Spearman correlation (aka. Spearman\'s *Ï*).

-   Just specify the method as spearman

-   X calculate the values between all actual values

    -   instead ranks both variables and calculate the correlation between ranked items

    -   e.g. All ranked highest in XXX have high correlation between ranked highest in YYY

```{r}
cor(d$life_expectancy, d$gdppc, use = "complete.obs", method = "spearman")
```

```{r}
cor.test(d$life_expectancy, d$gdppc, use = "complete.obs", method = "spearman")
```

## Quant x3

-   Correlation Matrix

-   Bubble Plot: 3rd variable visualized with size

-   Make it colorful: 3rd variable visualized with color

```{r}
library(GGally)
d |>
  select(gdppc, life_expectancy, population) |>
  ggpairs()
```

```{r}
# correlation matrix
d |> select(gdppc, life_expectancy, population) |>
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "Population"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")), # Upper triangle: Spearman correlation
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), # Diagnal: histogram
    lower = list(continuous = wrap("smooth", alpha = 0.1, color = "gray"))) # Lower triag: + linear trend line
```

```{r}
# correlation matrix
d |> select(gdppc, life_expectancy, population) |>
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "Population"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")), # Upper triangle: Spearman correlation
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), # Diagnal: histogram
    lower = list(continuous = wrap("smooth_loess", alpha = 0.1, color = "gray"))) # Lower triag: + loess trend line
```

### Transformation

-   If the variable is skewed, consider transformation to better visualise the data

```{r}
summary(d$population) # The population variable is skewed. Consider transofmration
```

```{r}
d |> select(gdppc, life_expectancy, population) |>
  mutate(population = log(population)) |> # log transformation
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "log(Population)"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")), # Upper triangle: Spearman correlation
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), # Diagnal: histogram
    lower = list(continuous = wrap("smooth_loess", alpha = 0.1, color = "gray"))) # Lower triag: + loess trend line
```

## Bubble Plot

```{r}
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population)) +
  labs(x = "GDP per capita", y = "Life Expectancy", title = "Wealth and Health in the World (2019)")
```

```{r}
# customized
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population), alpha = 0.3) +
  labs(x = "GDP per capita", y = "Life Expectancy", size = "Population",
       title = "Wealth and Health in the World (2019)")
```

```{r}
# annotated
d_anno_pop <- d |> filter(year == 2019) |> slice_max(order_by = population, n = 10)

d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population), alpha = 0.3) +
  geom_text(data = d_anno_pop, aes(x = gdppc, y = life_expectancy, label = country_text_id, size = population)) +
  labs(x = "GDP per capita", y = "Life Expectancy", size = "Population",
       title = "Wealth and Health in the World (2019)",
       caption = "Note: World's top 10 most populated countries are annotated.") +
  theme(legend.position = "none") # Remove the legend
```

```{r}
# make it colourful
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population, color = population), alpha = 0.3) +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       size = "Population", color = "Population", fill = "Population",
       title = "Wealth and Health in the World (2019)")
```

```{r}
# make it colourful x2
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population, color = population), alpha = 0.3) +
  scale_color_viridis_c(option = "A", trans = "log", direction = -1) +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       size = "Population", color = "Population", fill = "Population",
       title = "Wealth and Health in the World (2019)")
```

## Quant. X 4+?

It is *technically* feasible to plot more than 3 variables in the same figure, but...

-   You don\'t want your figure to be too \"busy\"

-   \"Less is more\" ðŸ˜¢

```{r}
# use colour to show infant mortality
d |> 
  filter(year == 2019) |>
  ggplot(aes(x = gdppc, y = life_expectancy)) +
  geom_point(aes(size = population, color = infant_mortality), alpha = 0.3) +
  scale_color_viridis_c(option = "A") +
  labs(x = "GDP per capita", y = "Life Expectancy", 
       size = "Population", color = "Infant Mortality",
       title = "Wealth and Health in the World (2019)")
```

```{r}
d |> filter(year == 2019) |>select(gdppc, life_expectancy, population, infant_mortality) |>
  mutate(population = log(population)) |>
  ggpairs(
    columnLabels = c("GDP per capita", "Life Expectancy", "log(Population)", "Infant Mortality"), # Label variables
    upper = list(continuous = wrap("cor", method = "spearman", color = "blue")),
    diag = list(continuous = wrap("barDiag", bins = 30, fill = "white", color = "black")), 
    lower = list(continuous = wrap("smooth_loess", alpha = 0.3, color = "gray")))
```
